---
title: "Minimum Viable Analysis"
author: "Aleca Borsuk"
date: "11/2/2020"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ape)
library(geiger)
library(ggtree)
library(knitr)
```

## Maximum Likelihood

Maximum likelihood tree for subset of Viburnum taxa constructed using chloroplast sequence data from NCBI and IQTree:

```{r}
# read in the phylogeny generated by IQtree 
phy = read.tree("concatenated.fasta.treefile")

# resolve multichotomies into a series of dichotomies with one (or more) branches of length zero
phy=multi2di(phy)

#root the phylogeny with V. clemensiae, sister to rest of genus
root(phy, phy[[4]][94])

#check if rooted
is.rooted(phy) 

#View the phylogeny
plot(phy, no.margin=TRUE,edge.width=1, cex = 0.5)

# Make ultrameric phylogeny (relative to root depth = 1)
phy_calibration <- makeChronosCalib(phy, node = "root", age.min = 1, age.max =
1, interactive = FALSE, soft.bounds = FALSE)

#make the phylogeny ultrameric using relative calibration
phy_ultrameric <- chronos(phy, lambda = 1, model = "correlated", calibration = phy_calibration, control = chronos.control() )
class( phy_ultrameric ) = "phylo"

#View the ultrameric phylogeny
plot(phy_ultrameric, no.margin=TRUE,edge.width=1, cex = 0.5)

```
## Pruning the Tree to Species of Interest
```{r}
# Read in species names of interest for comparative analysis. This is also the table with trait data
# Trait data is a combination of measured and dummy data (study in progress)
df <- read.csv("PalisadeTraits.csv", header = TRUE)

# Remove the species of interest from the list of species to drop, then drop tips
sp_keep <- df$Species
sp_full <- phy[[4]]
to.drop <- sp_full [! sp_full %in% sp_keep]

phy_reduced = drop.tip( phy_ultrameric, to.drop )

plot( phy_reduced )

```

## Mapping Character Data onto the Tree

Visualization of the character 'palisade.type' mapped onto the tree tips. 

```{r}

# View data in table format
kable(df)

# Plot tree with tip labels by palisade type
cols <- c("skyblue2","blue","gold","darkgoldenrod3")

palisade.type_plot = ggtree(phy_reduced) %<+% df + 
  geom_tiplab( fontface = "italic", offset = 0.005) +
  scale_color_manual(values=cols) + 
  xlim(0, .35) +
  #theme_tree2(legend.position='right')+
  geom_tippoint( aes(color=PalisadeType), size=3, alpha=1 )

palisade.type_plot

```

## Phylogenetically Informed ANOVA

Here using the tree from the above analysis to do a test run of a phylogenetically informed ANOVA. ANOVA method conducted using geiger v2.0 (Pennell et al. 2014):

```{r}
# Data formatting
# format discrete data as factor
df$PalisadeType <- as.factor(df$PalisadeType)
# anova function takes named vectors as input
df1= df[,2]
df2 = df[,3]
names(df1)=df[,1]
names(df2)=df[,1]

# run ANOVA
x1=aov.phylo(df2~df1, phy_reduced, nsim=50)
# formatted summary table
kable(print(attributes(x1)$summary))

# boxplot visualization
plot(df$Amax~df$PalisadeType, col=cols)
```
I believe the ANOVA test shows that there is a strong phylogenetic signal for maximum photosynthetic rate (Amax), but that it is not related to palisade cell type. 
