---
title: "Minimum Viable Analysis"
author: "Aleca Borsuk"
date: "11/2/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ape)
library(geiger)
```

## Maximum Likelihood

Maximum likelihood tree for subset of Viburnum taxa constructed using chloroplast sequence data from NCBI and IQTree:

```{r phy}
# read in the phylogeny generated by IQtree 
phy = read.tree("viburnum_alignment.fasta.treefile")
# root the tree; need to follow up to understand what this function is doing
phy=multi2di(phy)
#View the phylogeny
plot(phy)
```

## Phylogenetically Informed ANOVA

Here using the tree from the above analysis to do a test run of a phylogenetically informed ANOVA. ANOVA method conducted using geiger v2.0 (Pennell et al. 2014):

```{r anova}
# import trait data
df <- read.csv("PalisadeTraits.csv", header = TRUE)
# format discrete data as factor
df$PalisadeType <- as.factor(df$PalisadeType)
# anova function takes named vectors as input
df1= df[,2]
df2 = df[,3]
names(df1)=df[,1]
names(df2)=df[,1]

# run ANOVA
x1=aov.phylo(df2~df1, phy, nsim=50)
#print(attributes(x1)$summary) # summary table

# boxplot visualization
plot(df$Amax~df$PalisadeType)
```
Currently working with a subset of the full dataset, so getting some errors related to missing taxa, which is fine for now. I believe the ANOVA test shows that there is a strong phylogenetic signal for maximum photosynthetic rate (Amax), but that it is not related to palisade cell type. Just testing the method on secondary data; will for the final project have updated trait data, including more traits, to analyze. 
